Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-11-24 rnni_convexity.tex
first ideas abstract + discussion

\begin{abstract}

\end{abstract}

@@

For providing further evidence that the Cluster Theorem holds in $\rnni$, we can show computationally that it holds for the $\rnni$ graph on small trees with up to six taxa.
We did this by computing subgraph of $\rnni$ only containing trees that share a cluster and compare distances within this graph with distances in the whole $\rnni$ graph \autocite{}.
\todo{LC: github link}
We computed these graphs according to the algorithm of \autocite{Gavryushkin2018-ol}, discussed in Section~\ref{section:algorithms}.

@@

>>>>

\begin{abstract}
Within this paper we consider a graph of ranked trees, that are rooted binary phylogenetic trees where internal nodes are ordered.
The ranking of these internal nodes can be seen as a discretisation of trees where internal nodes are assigned divergence times.
For phylogenetic reconstruction such as Markov Chain Monte Carlo (MCMC) methods, it is important to understand the geometry of tree spaces.
This motivates our work, which focusses on understanding the geometry of the space of ranked trees, the so-called ranked nearest neighbour interchange ($\rnni$) graph.
Moreover, we are interested in understanding the complexity of computing distances in this graph.
We will see that even though there are similarities between the $\rnni$ graph on ranked trees and the classical nearest neighbour interchange ($\nni$) graph on unranked phylogenetic trees, some significant differences between these tree spaces suggest that there is a difference in the complexity of computing distances.
\end{abstract}

@@

For providing further evidence that the Cluster Theorem holds in $\rnni$, we can show computationally that it holds for the $\rnni$ graph on small trees with up to six taxa.
We did this by computing subgraph of $\rnni$ only containing trees that share a cluster and compare distances within this graph with distances in the whole $\rnni$ graph \autocite{}.
\todo{LC: github link}
We computed these graphs according to the algorithm of \autocite{Gavryushkin2018-ol} as discussed in Section~\ref{section:algorithms}.

@@

Within this paper we established some geometric properties of the $\rnni$ graph.
We furthermore pointed out some differences between this graph on ranked trees and the well studied $\nni$ graph on unranked trees.
This is of particular interest for analysing the complexity of computing distances in $\rnni$ as this problem is known to be $\np$-complete in $\nni$.
Our finding that the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex}) suggests that there is a difference in complexity between these two tree spaces.
This leaves hope that computing distances in $\rnni$ is, contrary to this problem in $\nni$, not $\np$-complete.
The question of how hard this problem actually is still remains open and will be subject of further research.

Our work in this paper highlights the importance of algorithms for computing paths in $\rnni$ in order to prove geometric properties such as diameter and radius.
One of the algorithms that has been introduced, $\findpath$, computed exact distances for all pairs of trees on up to six taxa.
For future investigations of the $\rnni$ space this algorithm might be a useful tool.
